Neural Connectome of the Ctenophore Statocyst

### 2025-05-16 manuscript/Manuscript.qmd
MS: edits, fig legends, refs

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than effector function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior. Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, and functional imaging revealed correlated activity patterns in balancer cilia. These results suggest that the aboral nerve net plays a central role in integrating sensory input and coordinating motor output.

@@

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, 620 of these images were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells with nuclei and cell bodies intact. Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

@@

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants.

![**Figure 1. Morphology of the aboral organ in *Mnemiopsis leidyi***\
(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the aboral organ (ao). Scale bar: 100 µm.\
(B) Schematic diagram of the aboral view of a cydippid larva. The two-tone coloration represents the biradial symmetry of the body. For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants. The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T). Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
(C) Aboral views of the aboral organ highlighting its spatial organization. The left panel presents a schematic representation of the aboral organ, illustrating the four body quadrants (Q1–Q4), color-coded consistently with panel (B). The right panel shows a corresponding differential interference contrast (DIC) image from the same perspective. The region enclosed by the dotted line delineates the boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows indicate the viewing directions corresponding to those planes. Scale bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\
(D) Lateral views of the aboral organ in two orthogonal planes captured using DIC microscopy. The left panel shows a lateral view in the sagittal (S) plane, while the right panel displays a view in the tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are enclosed within the dome (do). Scale bar: 50 µm. A–O indicates the aboral–oral body axis.\
(E) Schematic representations of the aboral organ in sagittal and tentacular planes. The left panel shows a lateral schematic in the sagittal plane (S), and the right panel shows the same in the tentacular plane (T), corresponding to the views in panel (D). Colors match the quadrant scheme shown in panels (B) and (C).\
(F) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
(G) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives. The left panel presents an aboral view of the aboral organ. The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view. Cells are color-coded according to their respective quadrants. Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines. Scale bar: 25 µm.](figures/Figure1.png)

@@

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynaptic structures, whereas ANN_Q3Q4 had 68 presynaptic structures.

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. Except for lithocytes, each cell type has a sample size of at least three cells. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within ANN Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.](figures/Figure2.png)

@@

The second group of cells that form synaptic contact with the ANN were the bridge cells. Bridge cells, first described by Tamm et al. in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells. Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells.

@@

In both The Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.

@@

![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
(A) Graph showing the number of synaptic outputs from the ANNs (ANN Q1-4: blue, ANN Q1Q2: orange, ANN Q3Q4: pink) circuit for each cell type. The balancer and bridge cell types receive relatively high outputs from the ANNs circuit. ANN represents the number of synapses within the ANNs circuit. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

In larvae with their sagittal plane facing the objective, we could compared balancer-cilia movements between Q1(4) and Q2(3). In other larvae oriented in the tentacular plane, we could simultaneously image Q1(3) and Q2(4). We used a high-speed camera and recorded at 100 fps for 2 minutes (12,000 frames). To analyse ciliary beating, we selected regions of interest (ROIs) in areas where brightness changes indicated ciliary beating. Ciliary beating was both manually quantified and plotted as kymographs.

@@

![**Figure 4. Differences in Balancer Cilia Movement Control Due to Variations in the Neural Circuit Pathways**\
(A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement. The microscope was tilted 90 degrees so that the stage was positioned vertically. A monochrome CMOS camera sensitive to near-infrared (NIR) light was used, synchronized with an 850 nm strobe light source. The movement of the balancer cilia was recorded using a 40× objective lens.\
(B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes. Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation.\
(C) Representative kymographs showing arrest and re-beat events of balancer cilia. Left and right balancer cilia were simultaneously recorded in either the sagittal (left) or tentacular plane (right). Arrowheads mark the onset of arrest (open) and re-beat (filled); horizontal arrows represent 1-second intervals.\
(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (gray boxes) between left and right balancer cilia. Colors of the box outlines indicate the imaging plane: sagittal (blue) or tentacular (orange). Each dot represents a single larva.\
(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from sagittal (left) and tentacular (right) planes. Positive and negative values represent opposite sides within each plane.\
(F) Mean rolling Pearson correlation of left–right balancer CBF calculated with a 20-frame window. Each dot represents an individual larva. Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, p = 0.014).](figures/Figure4.png)

@@

### Functional Nerve Net for Ciliary Coordination

In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy. Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

>>>>

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than sensory-motor function for the nerve net and inform our understanding of the diversity of nervous-system organisation across animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early genomic and single-cell transcriptomic studies failed to identify conserved neuronal markers in ctenophores [@Moroz2014; @Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were molecularly identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour. Clarifying this relationship between neural architecture and behavioural control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer-cell output [@Lowe1997].

To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cell. By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ. Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.

@@

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells that had intact nuclei and cell bodies. Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

@@

Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants. <!-- check number of cells, 927 in figure -->

![**Figure 1. Morphology of the aboral organ in *Mnemiopsis leidyi***\
(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the aboral organ (ao). Scale bar: 100 µm.\
(B) Schematic diagram of the aboral view of a cydippid larva. The two-tone coloration represents the biradial symmetry of the body. For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants. The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T). Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
(C) Aboral views of the aboral organ highlighting its spatial organization. The left panel presents a schematic representation of the aboral organ, illustrating the four body quadrants (Q1–Q4), color-coded consistently with panel (B). The right panel shows a corresponding differential interference contrast (DIC) image from the same perspective. The region enclosed by the dotted line delineates the boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows indicate the viewing directions corresponding to those planes. Scale bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\
(D) Lateral views of the aboral organ in two orthogonal planes captured using DIC microscopy. The left panel shows a lateral view in the sagittal (S) plane, while the right panel displays a view in the tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are enclosed within the dome (do). Scale bar: 50 µm. A–O indicates the aboral–oral body axis.\
(E) Schematic representations of the aboral organ in sagittal and tentacular planes. The left panel shows a lateral schematic in the sagittal plane (S), and the right panel shows the same in the tentacular plane (T), corresponding to the views in panel (D). Colors match the quadrant scheme shown in panels (B) and (C).\
(F) Example of cell tracing in CATMAID. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
(G) Morphological rendering of cells composing the aboral organ displayed in aboral view (left pannel), sagittal plane view (middle panel) and tentacular plane view (right panel). Cells are color-coded according to quadrants. Lithocytes (li) are represented as three gray spheres, balancers (bal) are depicted as gray lines. Scale bar: 25 µm.](figures/Figure1.png)

@@

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynaptic structures, whereas ANN_Q3Q4 had 68 presynaptic structures.

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) Morphological rendering of the large aboral nerve-net neuron ANN_Q1-4 spanning all four quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel). Spheres indicate the positions of nuclei.\
(B) Morphological rendering of the two smaller ANN neurons ANN_Q1Q2 (pink) and ANN_Q3Q4 (orange), each spanning two quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel).  Spheres indicate the positions of nuclei.\
(C) Number of mitochondria per cell for each cell type. The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey. N > 2 cells for each cell type except lithocytes. Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
(D) Position of mitochondria within ANN_Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles.\
(F) Position of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN_Q1-4.](figures/Figure2.png)

@@

The second group of cells that formed synaptic contacts with the ANN were the bridge cells. Bridge cells, first described in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells. Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells.

@@

In both the Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.

@@

![**Figure 3. Synaptic connectivity of the balancer and and bridge cells with the ANN.**\
(A) Number of synaptic inputs from the ANNs (ANN_Q1-4: blue, ANN_Q1Q2: orange, ANN_Q3Q4: pink) to each cell type. Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
<!-- check order of panels, change here or change figure -->
(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
(C) Morphological rendering of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

In larvae with their sagittal plane facing the objective, we could compare balancer-cilia movements between Q1(4) and Q2(3). In other larvae oriented in the tentacular plane, we could simultaneously image Q1(3) and Q2(4). We used a high-speed camera and recorded at 100 fps for 2 minutes (12,000 frames). To analyse ciliary beating, we selected regions of interest (ROIs) in areas where brightness changes indicated ciliary beating. Ciliary beating was both manually quantified and plotted as kymographs.

@@

![**Figure 4. Analysis of ciliary beating, arrests and re-beat across balancers in the four quadrants**\
(A) Schematic diagram of the differential interference contrast (DIC) microscopy setup used to image the movement of balancer cilia. The microscope was tilted 90 degrees so that the stage was positioned vertically. We used a 40x objective lens and a monochrome CMOS camera sensitive to near-infrared (NIR) light that was synchronized with an 850 nm strobe light source.\
(B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes. Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation.\
(C) Representative kymographs showing arrest and re-beat events of balancer cilia. Left and right balancer cilia were simultaneously recorded in either the sagittal (left) or tentacular plane (right). Arrowheads mark the onset of arrest (open) and re-beat (filled); horizontal arrows represent 1-second intervals.\
(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (grey boxes) between left and right balancer cilia. Colors of the box outlines indicate the imaging plane: sagittal (blue) or tentacular (orange). Each dot represents a single larva.\
(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from the sagittal (left) or the tentacular (right) planes. Positive and negative values represent opposite sides within each plane.\
(F) Mean rolling Pearson correlation of left–right balancer CBF calculated with a 20-frame window. Each dot represents an individual larva. Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, p = 0.014).](figures/Figure4.png)

@@

### An Aboral Nerve Net for Ciliary Coordination

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve net (ANNs)—with multiple nuclei and a morphology distinct from thesubepithelial nerve net (SNN) with a “beads-on-a-string” organization reported in one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.

@@

How do differences in balancer cilia coordination across anatomical planes shape behaviour? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.
